\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-09 papers/esop2010/main.tex
go

Consider the following first-order program\footnote{Throughout the document,
we use \ocaml~syntax for programs.} that accumulates the sum of the contents
of some array {\tt a} in some global reference {\tt c}, using an integer {\tt
i} as a counter:
\begin{ocaml}
  let sum () = 
    for i = 0 to Array.length a - 1 do
      c := !c + a.(i)
    done
\end{ocaml}
This simple program can be annotated in a standard Hoare logic (for
example the system of \cite{Filliatre00a}) as follows:
\begin{who}
  let sum () =
    { !c = 0 }
    for i = 0 to Array.length a - 1 do
      { !c = {{{ $\sum_0^{i-1}$ }}}  a.(i) }
      c := !c + a.(i)
    done
    { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}}  a.(i) }
\end{who}
Our program is a {\tt for} loop, so it requires an {\em invariant
annotation}, indicating which relation is preserved by the loop body. The
equalities that need to be proved are very simple. Also, it is obvious that
the program only modifies the reference {\tt c}, and this fact can be
formally confirmed by an effect analysis.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Two-state
postconditions are a standard way to get rid of most auxiliary variables.
However, the pre- and postcondition of {\tt f} may not refer to arbitrary
parts of the state, but only the part {\tt f} reads or modifies.  With these
conventions, one can now write a generic implementation and specification for
the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
of our prototype \who. There are a few syntactic differences with \ocaml, in
particular for arrays: the type of arrays is {\tt 'a farray ref}, the length
of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
the logic, and similarly access to the {\tt i}th element is {\tt get i !a} in
the program and {\tt get i !!a} in the logic.

@@

The store types $\efft{ε}$ describe a portion of the store and are basically
mappings from region variables to values. The domain of the mapping is the
effect $ε$, where effect variables stand for unknown parts of the domain. To
deal with objects of these types, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.

>>>>

Consider the following first-order program\footnote{Throughout the document,
we use \ocaml~syntax for programs.} that accumulates the sum of the contents
of some array {\tt a} in some global reference {\tt c}, using an integer {\tt
i} as a counter:
\begin{ocaml}
  let sum () = 
    for i = 0 to Array.length a - 1 do
      c := !c + a.(i)
    done
\end{ocaml}
This simple program can be annotated in a standard Hoare logic (for
example the Why system~\cite{Filliatre00a}) as follows:
\begin{who}
  let sum () =
    { !c = 0 }
    for i = 0 to Array.length a - 1 do
      { !c = {{{ $\sum_0^{i-1}$ }}}  a.(i) }
      c := !c + a.(i)
    done
    { !c = {{{ $\sum_0^{ {\tt Array.length~a} - 1}$ }}}  a.(i) }
\end{who}
Our program is a {\tt for} loop, so it requires an {\em invariant
annotation}, indicating which relation is preserved by the loop body. The
equalities that need to be proved are very simple. Also, it is obvious that
the program only modifies the reference {\tt c}, and this fact can be
formally confirmed by an effect analysis.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
Why~\cite{Filliatre00a} with the more powerful effect system of Talpin and
Jouvelot~\cite{talpin94taed} and the more powerful annotations of
Régis-Gianas/Pottier~\cite{regis-gianas-pottier-08} to obtain a modular
system for proving higher-order programs with side effects. As
in~\cite{regis-gianas-pottier-08}, we use operators \preml\ and \postml\ to
access the pre- and postcondition of a function. However, in our system, a
precondition takes as an argument not only the function argument, but also the
initial state, before the execution of the function. The postcondition takes
the argument and return value, but also the states before {\em and} after the
function call. Two-state postconditions are a standard way to get rid of most
auxiliary variables.  However, the pre- and postcondition of {\tt f} may not
refer to arbitrary parts of the state, but only the part {\tt f} reads or
modifies.  With these conventions, one can now write a generic implementation
and specification for the {\tt iter} function. It is
given~Fig.~\ref{fig:iter}, in the input syntax of our prototype \who. There
are a few syntactic differences with \ocaml, in particular for arrays: the
type of arrays is {\tt 'a farray ref}, the length of the array {\tt a} is {\tt
length !a} in the program and {\tt length !!a} in the logic, and similarly
access to the {\tt i}th element is {\tt get i !a} in the program and {\tt get
i !!a} in the logic.

@@

The store types $\efft{ε}$ describe a portion of the store and are basically
mappings from references to values. The domain of the mapping is the
effect $ε$, where effect variables stand for unknown parts of the domain. To
deal with objects of these types, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.
